Forecasting the 2024 US Presidential Election

### 2024-11-04 paper/paper.qmd
cleaned data

@@

# Filter poll_data to include only the specified states
poll_summary <- poll_data %>%
  filter(state %in% c("Michigan", "Georgia", "Nevada", "North Carolina", 
                      "New Hampshire", "Wisconsin", "Pennsylvania", "Arizona")) %>%
  group_by(state) %>%   
  summarise(
    dem_total_pct = sum(democrat_pct, na.rm = TRUE),
    rep_total_pct = sum(republican_pct, na.rm = TRUE),
    highly_educated_pct = mean(`What is the highest level of education you have attained? College graduate`, na.rm = TRUE),
    lower_educated_pct = mean(`What is the highest level of education you have attained? High school or less`, na.rm = TRUE),
    female_pct = mean(`Can you please tell me your gender? Female`, na.rm = TRUE),
    male_pct = mean(`Can you please tell me your gender? Male`, na.rm = TRUE),
    non_binary_other_pct = mean(`Can you please tell me your gender? Nonbinary or other`, na.rm = TRUE),
    caucasian_pct = mean(`For statistical purposes only, can you please tell me your ethnicity? White or Caucasian`, na.rm = TRUE),
    minority_pct = 100 - caucasian_pct,
    past_vote_biden = mean(`Who did you vote for in the 2020 election? Joe Biden`, na.rm = TRUE),
    past_vote_trump = mean(`Who did you vote for in the 2020 election? Donald Trump`, na.rm = TRUE)
  ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%  # Replace NA values with 0
  mutate(across(where(is.numeric), ~ round(., 1)))  # Round to the nearest tenth

@@

```{r}
#| label: fig-party-extremes
#| fig-cap: Demographic Differences Between Maryland and South Dakota
#| echo: false
#| warning: false
#| message: false

# Prepare the poll_summary data frame
poll_summary <- poll_data %>%
  mutate(party_diff = democrat_pct - republican_pct) %>%
  select(-c(2, 3, 4, 6, 7, 8, 9, 11)) %>%  # Remove specific columns by index
  select(-contains('Mode')) %>%             # Remove columns containing 'Mode' in their names
  select(-contains('registration'))

@@

  top_diff_vars %>%
    pivot_longer(
      cols = c("Blue_Mean", "Red_Mean"), 
      names_to = "Group", 
      values_to = "Mean"
    ) %>%
    mutate(Group = recode(Group, "Blue_Mean" = "Maryland", "Red_Mean" = "South Dakota"))
}

# Function to create the comparison plot
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 3
    ) +
    # Update the colors and labels for the groups
    scale_fill_manual(values = c("Maryland" = "blue", "South Dakota" = "red")) +
    labs(
      title = "Demographic Differences Between Maryland and South Dakota",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 4)
    )
}

# Main analysis function
analyze_state_differences <- function(data, blue_state, red_state, top_n_vars = 4) {
  diff_df <- compute_mean_differences(data, blue_state, red_state)

@@

# Run the analysis for Maryland and South Dakota
analysis_results <- analyze_state_differences(poll_summary, blue_state = "Maryland", red_state = "South Dakota", top_n_vars = 5)

# Display the plot
analysis_results$plot

```
Beyond identifying swing states, it's also important to understand the differences between the strongest Democratic ("blue") and Republican ("red") states. The bar graph in @fig-party-extremes highlights these differences. It focuses on key factors that create the biggest contrasts between Democratic-leaning and Republican-leaning states, helping to show how these factors differ between the most extreme red and blue states.

@@

# Calculate poll_summary with rounding to the nearest tenth
poll_summary <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    female_pct = round(mean(.data[[column_refs$female]], na.rm = TRUE), 1),
    male_pct = round(mean(.data[[column_refs$male]], na.rm = TRUE), 1),
    lean_trump_pct = round(mean(.data[[column_refs$lean_trump]], na.rm = TRUE), 1),
    lean_kamala_pct = round(mean(.data[[column_refs$lean_Kamala]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data <- poll_summary %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Female = female_pct,
    Male = male_pct,
    LeanTrump = lean_trump_pct,
    LeanKamala = lean_kamala_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct
  )

@@

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix_2, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Voting Preferences and State Lean",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1)  # Axis labels size
  )

@@

# Model {#sec-model}
## Model Development

### Forecasting Vote Outcomes in Key Swing States

@@

### Model Design and Approach

@@

### Key Components of Our Models

@@

### Assumptions Underlying Our Models

@@

### Bayesian Modeling Implementation

@@

### Results and Interpretation

@@

# Generate predictions for Democratic and Republican models
#predictions_dem <- posterior_predict(bayesian_model_dem, newdata = swing_data)
#predictions_rep <- posterior_predict(bayesian_model_rep, newdata = swing_data)

# Function to summarize predictions by state
#aggregate_summary <- function(predictions, state_labels) {
  #summary_df <- data.frame(State = character(), mean = numeric(), lower = numeric(), upper = numeric())
  #for (state in unique(state_labels)) {
    #state_indices <- which(state_labels == state)
    #state_predictions <- predictions[, state_indices]
    #state_mean <- mean(state_predictions)
    #state_lower <- quantile(state_predictions, 0.025)
    #state_upper <- quantile(state_predictions, 0.975)
    #summary_df <- rbind(summary_df, data.frame(State = state, mean = state_mean, lower = state_lower, upper = state_upper))
  #}
  #return(summary_df)
#}

# Summarize predictions for both Democrat and Republican models
#dem_summary <- aggregate_summary(predictions_dem, swing_data$State)
#rep_summary <- aggregate_summary(predictions_rep, swing_data$State)

# Combine summaries into the final prediction summary
#prediction_summary <- data.frame(
  #State = dem_summary$State,
  #DEM_Mean = dem_summary$mean,
  #DEM_Lower = dem_summary$lower,
  #DEM_Upper = dem_summary$upper,
  #REP_Mean = rep_summary$mean,
  #REP_Lower = rep_summary$lower,
  #REP_Upper = rep_summary$upper
#)

@@

# Modify the prediction summary
#prediction_summary <- prediction_summary %>%
  #select(State, DEM_Mean, REP_Mean) %>%  # Keep only the mean columns
  #mutate(Winner = ifelse(DEM_Mean > REP_Mean, "Democrat", "Republican"))

# Create a formatted table
#kable(
  #prediction_summary,
  #col.names = c("State", "Democrat Points", "Republican Points", "Winner"),
  #format = "latex", 
  #digits = 2,
  #escape = FALSE
#) %>%
  #kable_styling(
    #latex_options = c("striped", "hold_position", "condensed"), 
    #full_width = FALSE, 
    #font_size = 10
#)

@@

## Additional Data & Model Details
Include any technical details on data cleaning, model diagnostics, and posterior checks.

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

# References

>>>>

poll_data <- poll_data %>%
  mutate(across(12:ncol(.), ~ round(as.numeric(.), 1)))

time_data <- time_data %>%
  mutate(across(5:ncol(.), ~ round(as.numeric(.), 1)))

@@

# Filter poll_data to include only the specified states
poll_summary <- poll_data %>%
  filter(state %in% c("Michigan", "Georgia", "Nevada", "North Carolina", 
                      "New Hampshire", "Wisconsin", "Pennsylvania", "Arizona")) %>%
  group_by(state) %>%   
  summarise(
    dem_total_pct = sum(democrat_pct, na.rm = TRUE),
    rep_total_pct = sum(republican_pct, na.rm = TRUE),
    highly_educated_pct = mean(`What is the highest level of education you have attained? College graduate`, na.rm = TRUE),
    lower_educated_pct = mean(`What is the highest level of education you have attained? High school or less`, na.rm = TRUE),
    female_pct = mean(`Can you please tell me your gender? Female`, na.rm = TRUE),
    male_pct = mean(`Can you please tell me your gender? Male`, na.rm = TRUE),
    non_binary_other_pct = mean(`Can you please tell me your gender? Nonbinary or other`, na.rm = TRUE),
    caucasian_pct = mean(`For statistical purposes only, can you please tell me your ethnicity? White or Caucasian`, na.rm = TRUE),
    minority_pct = 100 - caucasian_pct,
    past_vote_biden = mean(`Who did you vote for in the 2020 election? Joe Biden`, na.rm = TRUE),
    past_vote_trump = mean(`Who did you vote for in the 2020 election? Donald Trump`, na.rm = TRUE)
  )

@@

```{r}
#| label: fig-party-extremes
#| fig-cap: Demographic Differences Between California and South Dakota
#| echo: false
#| warning: false
#| message: false

# Prepare the poll_summary data frame
poll_summary <- poll_data %>%
  mutate(party_diff = democrat_pct - republican_pct) %>%
  select(-c(2, 3, 4, 6, 7, 8, 9, 11)) %>%  # Remove specific columns by index
  select(-contains('Mode')) %>%             # Remove columns containing 'Mode' in their names
  select(-contains('registration')) %>%
  select(-democrat_pct, -republican_pct)    # Exclude 'democrat_pct' and 'republican_pct'

@@

  top_diff_vars %>%
    pivot_longer(
      cols = c("Blue_Mean", "Red_Mean"), 
      names_to = "Group", 
      values_to = "Mean"
    ) %>%
    mutate(Group = recode(Group, "Blue_Mean" = "California", "Red_Mean" = "South Dakota"))
}

# Function to create the comparison plot
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 3
    ) +
    # Update the colors and labels for the groups
    scale_fill_manual(values = c("California" = "blue", "South Dakota" = "red")) +
    labs(
      title = "Demographic Differences Between California and South Dakota",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 4)
    )
}

# Main analysis function
analyze_state_differences <- function(data, blue_state, red_state, top_n_vars = 5) {
  diff_df <- compute_mean_differences(data, blue_state, red_state)

@@

# Run the analysis for California and South Dakota
analysis_results <- analyze_state_differences(poll_summary, blue_state = "California", red_state = "South Dakota", top_n_vars = 6)

# Display the plot
analysis_results$plot
```
Beyond identifying swing states, it's also important to understand the differences between the strongest Democratic ("blue") and Republican ("red") states. The bar graph in @fig-party-extremes highlights these differences. It focuses on key factors that create the biggest contrasts between Democratic-leaning and Republican-leaning states, helping to show how these factors differ between the most extreme red and blue states.

@@

# Calculate poll_summary with rounding to the nearest tenth
poll_summary <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    female_pct = round(mean(.data[[column_refs$female]], na.rm = TRUE), 1),
    male_pct = round(mean(.data[[column_refs$male]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding lean_trump and lean_kamala
corr_data <- poll_summary %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Female = female_pct,
    Male = male_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct
  )

@@

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix_2, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Voting Preferences and State Lean",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 10, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1)  # Axis labels size
  )

@@

# Model {#sec-model}

## Forecasting Vote Outcomes in Key Swing States

@@

## Model Design and Approach

@@

## Key Components of Our Models

@@

## Assumptions Underlying Our Models

@@

## Bayesian Modeling Implementation

@@

## Results and Interpretation

@@

# Generate predictions for Democratic and Republican models
predictions_dem <- posterior_predict(bayesian_model_dem, newdata = swing_data)
predictions_rep <- posterior_predict(bayesian_model_rep, newdata = swing_data)

# Function to summarize predictions by state
aggregate_summary <- function(predictions, state_labels) {
  summary_df <- data.frame(State = character(), mean = numeric(), lower = numeric(), upper = numeric())
  for (state in unique(state_labels)) {
    state_indices <- which(state_labels == state)
    state_predictions <- predictions[, state_indices]
    state_mean <- mean(state_predictions)
    state_lower <- quantile(state_predictions, 0.025)
    state_upper <- quantile(state_predictions, 0.975)
    summary_df <- rbind(summary_df, data.frame(State = state, mean = state_mean, lower = state_lower, upper = state_upper))
  }
  return(summary_df)
}

# Summarize predictions for both Democrat and Republican models
dem_summary <- aggregate_summary(predictions_dem, swing_data$State)
rep_summary <- aggregate_summary(predictions_rep, swing_data$State)

# Combine summaries into the final prediction summary
prediction_summary <- data.frame(
  State = dem_summary$State,
  DEM_Mean = dem_summary$mean,
  DEM_Lower = dem_summary$lower,
  DEM_Upper = dem_summary$upper,
  REP_Mean = rep_summary$mean,
  REP_Lower = rep_summary$lower,
  REP_Upper = rep_summary$upper
)
```

@@

# Modify the prediction summary
prediction_summary <- prediction_summary %>%
  select(State, DEM_Mean, REP_Mean) %>%  # Keep only the mean columns
  mutate(Winner = ifelse(DEM_Mean > REP_Mean, "Democrat", "Republican"))

# Create a formatted table
kable(
  prediction_summary,
  col.names = c("State", "Democrat Points", "Republican Points", "Winner"),
  format = "latex", 
  digits = 2,
  escape = FALSE
) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "condensed"), 
    full_width = FALSE, 
    font_size = 10
  )

@@

# Additional data details
Include any technical details on data cleaning, model diagnostics, and posterior checks.
# References
